Denotational Semantics for Gradual Typing in Synthetic Guarded Domain Theory

### 2023-04-25 paper-new/paper.tex
work on paper (changes to intro, work on description of syntax)

\begin{abstract}
    We develop a denotational semantics for a gradually typed language
    with effects that is adequate and proves the graduality theorem.
    %
    The denotational semantics is constructed using \emph{synthetic
    guarded domain theory} working in a type theory with a later
    modality and clock quantification.
    %
    This provides a remarkably simple presentation of the semantics,
    where gradual types are interpreted as ordinary types in our ambient
    type theory equipped with an ordinary preorder structure to model
    the error ordering.
    %
    This avoids the complexities of classical domain-theoretic models
    (New and Licata) or logical relations models using explicit
    step-indexing (New and Ahmed).
    %
    In particular, we avoid a major technical complexity of New and
    Ahmed that requires two logical relations to prove the graduality
    theorem.

@@

  \section{Introduction}

  % gradual typing, graduality
  \subsection{Gradual Typing and Graduality}
  Gradual typing allows a language to have both statically-typed and dynamically-typed terms;
  the statically-typed terms are type checked at compile time, while type checking for the 
  dynamically-typed terms is deferred to runtime.

  Gradually-typed languages support the smooth migration from dynamic typing
  to static typing, in that the programmer can initially leave off the
  typing annotations and provide them later without altering the meaning of the
  program. Moreover, a key feature of gradually typed languages is that the interaction
  between the static and dynamic components of the codebase should be safe -- i.e., should
  preserve the guarantees made by the static types.
  In other words, in the static portions of the codebase, type soundness must be preserved.

  In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
  the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
  forms of the language (e.g., pattern matching, field reference, etc.).
  Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
  type checking is made explicit through the insertion of \emph{type casts}.

  ...

  % Up and down casts
  There is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that $A$ is a \emph{more
  precise} type than $B$.
  There a dynamic type $\dyn$ with the property that $A \ltdyn \dyn$ for all $A$.

  If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
  and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
  Upcasts always succeed, while downcasts may fail at runtime.

  % Graduality property
  A key property that any gradually-typed language should satisfy is the \emph{dynamic gradual guarantee},
  originally defined by Siek, Vitousek, Cimini, and Boyland \cite{TODO}.
  New et al refer to this property as \emph{graduality}, by analogy with parametricity.
  Intuitively, graduality says that in going from a dynamic to static style
  should not introduce changes in the meaning of the program.
  More specifically, making the types more precise (replacing $\dyn$ with a more specific type
  such as integers) will either result in the same behavior as the less precise program,
  or cause a dynamic type error.

  % Observational equivalence and approximation
  % TODO

  \subsection{Prior Approach}
  % Proving graduality via LR
  The usual approach to proving graduality uses the technique of \emph{logical relations}.
  Specifially, a logical relation is constructed and showed to be sound with respect to
  observational approximation. Because the dynamic type is modeled as a recursive type,
  the logical relation needs to be paramterized by natural numbers, a so-called
  step-indexed logical relation. Reasoning about step-indexed logical relations
  is tedious and error-prone, and there are several very subtle aspects that must
  be taken into account in the proofs.

  % Difficulties in prior semantics
  % TODO

  % - two separate logical relations for expressions
  % - transitivity

  % synthetic guarded domain theory, denotational semantics therein
  \subsection{Synthetic Guarded Domain Theory}
  One way to avoid the tedious reasoning associated with step-indexing is to work
  axiomatically inside of a logical system that can reason about non-well-founded recursive
  constructions while abstracting away the specific details of step-indexing required
  if we were working analytically.
  The system that proves useful fot this purpose is called \emph{synthetic guarded
  domain theory}, or SGDT for short. We provide a brief overview here, but more
  details can be found in \cite{TODO}.

  In SGDT, there is the notion of a \emph{clock}, an abstract type that tracks the
  passage of time through \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
  as evidence that one unit of time has passed according to the clock $k$.

  SGDT also introduces a modality $\later : \type \to \type$, (pronounced ``later'').
  Given a clock $k$ and type $T$, the type $\later_k T$ represents a value of type
  $T$ one unit of time in the future according to clock $k$.
  The type $\later T$ is represented as a function from ticks of a clock $k$ to $T$.
  There is an axiom called $\lob$-induction which is stated as follows:

  \[
    \lob : \forall T, (\later T \to T) \to T.
  \]

  That is, to construct a term of type $T$, it suffices to assume that we have access to
  such a term ``later'' and use that to help us build a term ``now''.

  In particular, this axiom applies to propositions $P : \texttt{Prop}$.

  \subsection{Overview of Remainder of Paper}
  In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
  for which we will prove graduality. Important here are the notions of syntactic
  type precision and term precision.

  In Section \ref{sec:domain-theory}, we define several fundamental constructions
  internal to SGDT that will be needed when we give a denotational semantics to
  the language. This includes the notion of Predomains as well as the concept
  of EP-Pairs.

  In Section \ref{Semantics}, we define the denotational semantics for the gradually-typed
  language

\section{GTLC}\label{sec:GTLC}

\subsection{Syntax}

\subsection{Type Precision}

\subsection{Term Precision}

\section{Domain-Theoretic Constructions}\label{sec:domain-theory}

\subsection{Predomains}

\subsection{Weak Bisimilarity Relation}

\subsection{EP-Pairs}

\section{Semantics}\label{sec:semantics}

\subsection{Types as Predomains}

\subsection{Terms as Monotone Functions}

\subsection{Type Precision as EP-Pairs}

\subsection{Term Precision}
% Homogeneous vs heterogeneous term precision

\section{Graduality}\label{sec:graduality}

@@

\subsection{Synthetic Ordering}

While the use of synthetic guarded domain theory allows us to very
conveniently work with non-well-founded recursive constructions while
abstracting away the precise details of step-indexing, we do work with
the error ordering in a mostly analytic fashion in that gradual types
are interpreted as sets equipped with an ordering relation, and all
terms must be proven to be monotone.
%
It is possible that a combination of synthetic guarded domain theory
with \emph{directed} type theory would allow for an a synthetic
treatment of the error ordering as well.

>>>>

\begin{abstract}
    We develop a denotational semantics for a simple gradually typed language
    that is adequate and proves the graduality theorem.
    %
    The denotational semantics is constructed using \emph{synthetic
    guarded domain theory} working in a type theory with a later
    modality and clock quantification.
    %
    This provides a remarkably simple presentation of the semantics,
    where gradual types are interpreted as ordinary types in our ambient
    type theory equipped with an ordinary preorder structure to model
    the error ordering.
    %
    This avoids the complexities of classical domain-theoretic models
    (New and Licata) or logical relations models using explicit
    step-indexing (New and Ahmed).
    %
    In particular, we avoid a major technical complexity of New and
    Ahmed that requires two logical relations to prove the graduality
    theorem.

@@

\input{intro}

\input{technical-background}

\input{syntax}

\input{semantics}

\input{graduality}

@@

% \subsection{Synthetic Ordering}

% While the use of synthetic guarded domain theory allows us to very
% conveniently work with non-well-founded recursive constructions while
% abstracting away the precise details of step-indexing, we do work with
% the error ordering in a mostly analytic fashion in that gradual types
% are interpreted as sets equipped with an ordering relation, and all
% terms must be proven to be monotone.
% %
% It is possible that a combination of synthetic guarded domain theory
% with \emph{directed} type theory would allow for an a synthetic
% treatment of the error ordering as well.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}
